---
title: "Causal Inference Assignment"
output: html_document
date: "2026-02-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message = FALSE}
library(dplyr)
library(dagitty)
library(ggplot2)
library(GGally)
library(gridExtra)
library(stdReg)
library(cobalt)
library(survey)
library(MatchIt) 
```

## Data preparation

```{r}
# Load the data
load("obesity.Rda")

obesity <- obesity |>
  mutate(diabetes = ifelse(diabetes == "No", 0, 1)) |>
  mutate(education = ifelse(education == "other", 0, 1)) |>
  mutate(income = ifelse(income == "low or normal", 0, 1)) |>
  mutate(current.smoking = ifelse(current.smoking == "no", 0, 2)) |>
  mutate(alcohol_level = case_when(alcohol == 0 ~ 0,
                                   alcohol > 0 & alcohol < 4 ~ 1,
                                 alcohol >= 4 ~ 2)) |>
  mutate(medstatin = ifelse(medstatin == "no", 0, 1)) |>
  mutate(medglucose = ifelse(medglucose == "no", 0, 1)) |> 
  # Remove columns that are not needed for this research
  select(-c("nr", "CVD", "MI", "angina", "stroke", "death")) 
```

We can combine some of the variables into 1 such as:

-   Mean arterial blood pressure = DIABP + (SYSBP - DIABP)/3
-   Social economic factors
    -   education: other (0), high (1)
    -   income: low or normal (0), high (1)
    -   We combine these into one factor
        -   (0): none of the factors are high
        -   (1): one of factors are high
        -   (2): both are high
-   Substance abuse
    -   alcohol: it ranges from 0 to 17 (do table(obesity\$alcohol) )
        -   0: no alcohol (0)
        -   1-3: occasionally drinker (1)
        -   4-17: heavy drinker (2)
    -   current.smoking: no (0), yes (2)
    -   We combine these into one factor, which is a sum of the alcohol and current.smoking levels
        -   (0): no smoking, no alcohol
        -   (1): only occasionally drinking
        -   (2): smoker or heavy drinker
        -   (3): smoking and occasionally drinker
        -   (4): smoking and heavy drinker
-   Medication:
    -   Use of medstatin and medglucose
    -   We combine these into one factor
        -   (0): none
        -   (1): one of them is used
        -   (2): both are used

```{r}
# Arterial blood pressure
obesity$ARTBP <- obesity$DIABP + (obesity$SYSBP - obesity$DIABP)/3

# Social economic factors
obesity$social_factors <- as.factor(obesity$income + obesity$education)

# Substance abuse 
obesity$substance_use <- as.factor(obesity$alcohol_level + obesity$current.smoking)

# Medication 
obesity$medication <- as.factor(obesity$medstatin + obesity$medglucose)

# Complete-cases only dataframe
obesity_na <- obesity |>
  select(-c("education", "income", "current.smoking", "alcohol", 
            "medstatin", "medglucose", "SYSBP", "DIABP")) |>
  filter(diabetes != "NA") |> # We only look at obesity
  na.omit() # Remove rows with NA value in any column
```

The obesity_na dataframe does not contain any NAs. We're using that one for performing the DAG testing, but later on we need to check the missingness of the obesity values and some of the covariates

## Descriptive Analysis

```{r}
# Diabetes counts in data
table(obesity$diabetes)

# Average values grouped by diabetes Yes/No
obesity |> 
  group_by(diabetes) |>
  summarize(mean_age = mean(age),
            sex_female = sum(sex == "female") / length(sex) * 100, # %
            medhyper_yes = sum(medhyper == "yes") / length(medhyper) * 100, # %
            bmi = mean(bmi),
            ARTBP = mean(ARTBP, na.rm = T))

# Medhyper takers 
sum(obesity$medhyper == "yes")
```

## DAG

```{r}
# Specify the DAG
g.diabetes <- dagitty('dag {
  social_factors[pos="-2, 1"]
  substance_use [pos="-2, 3"]
  sex           [pos="0, -2"]
  medhyper      [exposure,pos="1, 0"]
  bmi           [pos="2, -2"]
  waist         [pos="0, -1"]
  age           [pos="1.5, 1.5"]
  ARTBP         [pos="0, 1.5"]
  diabetes      [outcome,pos="-2, -2"]
  medication    [pos="2, 3"]
  social_factors -> substance_use
  substance_use -> {ARTBP}
  ARTBP -> diabetes
  sex -> {diabetes bmi}
  medhyper -> {ARTBP}
  bmi -> {waist medhyper medication}
  waist -> {diabetes ARTBP}
  age -> {bmi ARTBP medhyper medication}
  medication -> {ARTBP}
  }
')

# Plot the DAG
plot(g.diabetes)
```

```{r}
# Perform local tests 
r <- localTests(g.diabetes, obesity_na, 
                type = "cis.pillai") 

# Adjusts for multiple testing with Holm-Bonferri correction
r <- r[p.adjust(r$p.value) < 0.05, ] # Retain only significant results

# Plot the conditional independencies inconsistent with data
r <- r[order(r$p.value), ] # Order by significance
plotLocalTestResults(r, bty = "n", 
                     xlim = c(0, 1), ylim = c(0.1, 3.1), 
                     axis.pars = list(las = 1, lty = 0, cex.axis = 0.8))

# Variables to account for 
adjustmentSets(g.diabetes, effect = "total") 
```

type = "cis.pillai" is used as our data consists of both continuous and categorical variables

## G-modeling

```{r}
# Outcome model with interactions -> Y = diabetes, X = medhyper, C = confounders
model_g <- glm(diabetes ~ medhyper * age + medhyper * bmi, 
               family = binomial(link = "logit"), 
               data = obesity_na)
summary(model_g)
```

```{r}
# Predictions 
fit_std <- stdGlm(fit = model_g, data = obesity_na, X = "medhyper")

# Estimates with 95% CIs
summary(fit_std)

# ATE (Risk Difference) and its 95% CI
summary(fit_std, contrast = "difference", reference = "no")
```

## Missing values

```{r}
# Check the rows which have missing values
res <- obesity[!complete.cases(obesity),]
res

# Percentage of observations with missing data 
(212/10277) * 100

# Amount of missing values per variable
missing <- res |> summarise(across(everything(), ~ sum(is.na(.)))) 
missing # Counts
(missing/10065) * 100 # Percentages
```

## Propensity score

### Reweighting

```{r}
# Fit propensity model
ps_fit <- glm(medhyper ~ age + bmi,
              family = "binomial", data = obesity_na)
summary(ps_fit)

# Calculate propensity
obesity_na$ps <- fitted.values(ps_fit)
summary(obesity_na$ps)

# Generate corresponding weights
obesity_na$ipw <- (obesity_na$medhyper == "yes")/obesity_na$ps +
  (obesity_na$medhyper == "no")/(1 - obesity_na$ps)
summary(obesity_na$ipw)

# Calculate truncated weights
ipw.p99 <-quantile(obesity_na$ipw, 0.99) # 99th percentile
obesity_na$ipw.trunc <- ifelse(obesity_na$ipw > ipw.p99, ipw.p99, obesity_na$ipw)
summary(obesity_na$ipw.trunc)
```

```{r}
# Define new dataframe to assess balance
vars <- c("age", "bmi")
covariates <- obesity_na[, vars] 

# Standardised differences between treated and untreated patients
bal.tab(covariates, treat = obesity_na$medhyper, weights = obesity_na$ipw.trunc, 
        method = "weighting", un = TRUE)
```

```{r}
# Love plot
pdf(file = "Loveplot_medhyper.pdf")
p1 <- love.plot(covariates, treat = obesity_na$medhyper, 
               weights = obesity_na$ipw.trunc, method = "weighting", 
               binary = "std", threshold = 0.1, title = "Covariate Balance")
p1 + 
  theme( # Increase text size 
    legend.key.size = unit(1, "line"),
    legend.position = "right",
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13),
    axis.title.x = element_text(size = 12),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold")) +
  guides(color = guide_legend(override.aes = list(size = 2.5)))

dev.off()
```

```{r}
# Histogram
pdf(file = "Propensity_medhyper.pdf")
par(mfrow=c(2,1))
hist(obesity_na$ps[obesity_na$medhyper == "yes"], xlim = c(0, 1),
     main = "Medhyper", xlab = "Propensity Score")


hist(obesity_na$ps[obesity_na$medhyper == "no"], xlim = c(0, 1),
     main = "No Medhyper", xlab = "Propensity Score")
dev.off()
```

The histograms show considerable overlap

```{r}
# Calculate ATE (with the truncated weights)
d.w.trunc <- svydesign(~1, weights = obesity_na$ipw.trunc, data = obesity_na)
fit.ipw.trunc <- svyglm(diabetes ~ medhyper, design = d.w.trunc)
summary(fit.ipw.trunc)

# Get the CI 
confint(fit.ipw.trunc)
```

Estimates ATE with truncated weights: 0.005. CI: [-0.004, 0.014]

### Matching

Out of the 10065 observations, 5742 are matched (2871 treated, and 2871 untreated)

```{r}
# Match with a caliper
match.out <- matchit(medhyper ~ age + bmi, data = obesity_na,
                      method = "nearest", distance = "glm", caliper = 0.1)
match.out
summary(match.out)

# Analyse the matches
match.data <- match.data(match.out)
table(match.data$medhyper, match.data$diabetes) 
prop.table(table(match.data$medhyper, match.data$diabetes) , margin=1)
```
