---
title: "Untitled"
output: html_document
date: "2026-02-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message = FALSE}
library(dplyr)
library(dagitty)
library(ggplot2)
library(GGally)
library(gridExtra)
```

## Data preparation

```{r}
# Load the data
load("obesity.Rda")

obesity <- obesity |>
  mutate(diabetes = ifelse(diabetes == "No", 0, 1)) |>
  select(-c("nr", "medstatin", "medglucose", "CVD", 
            "MI", "angina", "stroke", "death")) 

obesity_na <- obesity |>
  filter(diabetes != "NA") |> # We only look at obesity
  na.omit() # Remove rows with NA value in any column
```

The obesity_na dataframe does not contain any NAs. I'm using that one for performing the DAG testing, but later on we need to take into the missingness of the obesity values and some of the covariates

```{r}
# Test correlations for blood pressure
cor.test(obesity$SYSBP, obesity$DIABP, method = "pearson")

# Test correlation for substance abuse

```

I tried "including" the correlations in our DAG, but not much has changed in during the localTests

The code to include the correlations in our DAG:

BP [latent, pos="-3, 1.5"]

substance.use [latent pos="-2, 3.5"]

substance.use -\> {alcohol current.smoking}

BP -\> {SYSBP DIABP}

## DAG

```{r}
# Specify the DAG
g.diabetes <- dagitty('dag {
  education     [pos="-1, -2"]
  income        [pos="-2.5, 0"]
  alcohol       [pos="-1, 3"]
  current.smoking [pos="-2.5, 3"]
  sex           [pos="1, -2"]
  medhyper      [pos="0, 3"]
  bmi           [pos="2, -1"]
  waist         [pos="1, 0"]
  age           [pos="1.5, 1.5"]
  SYSBP         [pos="-0.5, 1"]
  DIABP         [pos="-1.5, 1.5"]
  diabetes      [outcome,pos="-0.5, -0.5"]

  education -> {income}
  income -> diabetes
  sex -> {diabetes bmi}
  medhyper -> {SYSBP DIABP}
  bmi -> waist
  waist -> diabetes
  age -> {bmi SYSBP DIABP}
  alcohol -> {SYSBP DIABP}
  SYSBP -> diabetes
  DIABP -> diabetes
  current.smoking -> {SYSBP DIABP}
  
  }
')

# Plot the DAG
plot(g.diabetes)
```

```{r}
# Perform local tests 
r <- localTests(g.diabetes, obesity_na, type = "cis.pillai")

# Adjusts for multiple testing with Holm-Bonferri correction
r <- r[p.adjust(r$p.value) < 0.05, ] # Retain only significant results

# Plot the conditional independencies inconsistent with data
#r <- r[order(r$p.value), ] # Order by significance
plotLocalTestResults(r, bty = "n", 
                     xlim = c(0, 1), ylim = c(0.1, 3.1), 
                     axis.pars = list(las = 1, lty = 0, cex.axis = 0.8))
```

type = "cis.pillai" is used as our data consists of both continuous and categorical variables

## G-modeling

```{r}
# 
lin_model <- lm(diabetes ~ medhyper*waist + medhyper*sex + medhyper*SYSBP 
                + medhyper*DIABP +  medhyper*alcohol + medhyper*current.smoking,
                data = obesity)

summary(lin_model)
```

## Missing values

```{r}
# Check the rows which have missing values
res <- obesity[!complete.cases(obesity),]
print(res)
```

We see that most of the missing data is on education, then on smoking.status and ...
