---
title: "Untitled"
output: html_document
date: "2026-02-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message = FALSE}
library(dplyr)
library(dagitty)
library(ggdag)
library(ggplot2)
library(GGally)
library(gridExtra)
```

## Data preparation

```{r}
# Load the data
load("obesity.Rda")

obesity <- obesity |>
  mutate(diabetes = ifelse(diabetes == "No", 0, 1)) |>
  mutate(education = ifelse(education == "other", 0, 1)) |>
  mutate(income = ifelse(income == "low or normal", 0, 1)) |>
  mutate(current.smoking = ifelse(current.smoking == "no", 0, 2)) |>
  mutate(alcohol_level = case_when(alcohol == 0 ~ 0,
                                   alcohol > 0 & alcohol < 4 ~ 1,
                                 alcohol >= 4 ~ 2)) |>
  mutate(medstatin = ifelse(medstatin == "no", 0, 1)) |>
  mutate(medglucose = ifelse(medglucose == "no", 0, 1)) |> 
  # Remove columns that are not needed for this research
  select(-c("nr", "CVD", "MI", "angina", "stroke", "death")) 


```

We can combine some of the variables into 1 such as:

-   Mean arterial blood pressure = DIABP + (SYSBP - DIABP)/3
-   Social economic factors
    -   education: other (0), high (1)
    -   income: low or normal (0), high (1)
    -   We combine these into one factor
        -   (0): none of the factors are high
        -   (1): one of factors are high
        -   (2): both are high
-   Substance abuse
    -   alcohol: it ranges from 0 to 17 (do table(obesity\$alcohol) )
        -   0: no alcohol (0)
        -   1-3: occasionally drinker (1)
        -   4-17: heavy drinker (2)
    -   current.smoking: no (0), yes (2)
    -   We combine these into one factor, which is a sum of the alcohol and current.smoking levels
        -   (0): no smoking, no alcohol
        -   (1): only occasionally drinking
        -   (2): smoker or heavy drinker
        -   (3): smoking and occasionally drinker
        -   (4): smoking and heavy drinker
-   Medication:
    -   Use of medstatin and medglucose
    -   We combine these into one factor
        -   (0): none
        -   (1): one of them is used
        -   (2): both are used

```{r}
# Arterial blood pressure
obesity$ARTBP <- obesity$DIABP + (obesity$SYSBP - obesity$DIABP)/3

# Social economic factors
obesity$social_factors <- as.factor(obesity$income + obesity$education)

# Substance abuse 
obesity$substance_use <- as.factor(obesity$alcohol_level + obesity$current.smoking)

# Medication 
obesity$medication <- as.factor(obesity$medstatin + obesity$medglucose)

# Create the dataframe for the DAG testing
obesity_na <- obesity |>
  select(-c("education", "income", "current.smoking", "alcohol", 
            "medstatin", "medglucose", "SYSBP", "DIABP")) |>
  filter(diabetes != "NA") |> # We only look at obesity
  na.omit() # Remove rows with NA value in any column
```

The obesity_na dataframe does not contain any NAs. I'm using that one for performing the DAG testing, but later on we need to take into the missingness of the obesity values and some of the covariates

## DAG

```{r}
# Specify the DAG
g.diabetes <- dagitty('dag {
  social_factors[pos="-2.5, 1"]
  substance_use [pos="-2.5, 3"]
  sex           [pos="1, -2"]
  medhyper      [exposure,pos="0, 3"]
  bmi           [pos="2, -1"]
  waist         [pos="1, 0"]
  age           [pos="1.5, 1.5"]
  ARTBP         [pos="-0.5, 1"]
  diabetes      [outcome,pos="-0.5, -1"]
  medication    [pos="-2, -1"]

  social_factors -> substance_use
  substance_use -> {ARTBP}
  ARTBP -> diabetes
  sex -> {diabetes bmi}
  medhyper -> {ARTBP}
  bmi -> {waist medhyper medication}
  waist -> {diabetes ARTBP}
  age -> {bmi ARTBP medhyper medication}
  medication -> {ARTBP}
  }
')

# Plot the DAG
plot(g.diabetes)
```

```{r}
# Perform local tests 
r <- localTests(g.diabetes, obesity_na, type = "cis.pillai")

# Adjusts for multiple testing with Holm-Bonferri correction
r <- r[p.adjust(r$p.value) < 0.05, ] # Retain only significant results

# Plot the conditional independencies inconsistent with data
#r <- r[order(r$p.value), ] # Order by significance
plotLocalTestResults(r, bty = "n", 
                     xlim = c(0, 1), ylim = c(0.1, 3.1), 
                     axis.pars = list(las = 1, lty = 0, cex.axis = 0.8))

# Variables to account for 
adjustmentSets(g.diabetes, effect = "total") 
```

type = "cis.pillai" is used as our data consists of both continuous and categorical variables

## G-modeling

```{r}

```

## Missing values

```{r}
# Check the rows which have missing values
res <- obesity[!complete.cases(obesity),]
print(res)
```

We see that most of the missing data is on education, then on smoking.status and ...
